A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2011-11-22 papers/contact/paper.tex
a bit of work on contact paper Tested everything.

@@

\begin{widetext}

@@

One reasonably simple question to ask is what the \emph{mean} contact
density of an inhomogeneous system is.  This comes down to simply
reproducing the derivation for a homogeneous system using the
FMT functional.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR}
\end{align}
So to compute this we need
\begin{align}
  \frac{d A_{HS}}{d R} &=
  \frac{\partial A_{HS}}{\partial R} \\
  &+
  \int \left(
  \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
  \frac{d n_3(\mathbf{r}')}{d R}
  +
  \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
  \frac{d n_2(\mathbf{r}')}{d R}
  + \cdots
  \right) d\mathbf{r}'
\end{align}

@@

\begin{align}
  \frac{d A_{HS}}{d R} &=
  \int \mathbf{dr}' \left(
    -\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3)
    + \cdots
  \right)
\end{align}

@@

When we consider an inhomogeneous system, we may ask for \emph{which}
sphere we are interested in computing the contact density.  One way to
answer this is to specify a sphere at a fixed position $\mathbf{r}$.
By making the hard-sphere radius dependent on position, we can compute
the pressure on the hard spheres located at $\mathbf{r}$.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{4\pi (2R)^2} \frac{\delta A_{HS}}{\delta R(\mathbf{r})}
\end{align}
So to compute this we need
\begin{align}
  \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
  \int \left(
  \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
  \frac{\delta n_3(\mathbf{r}')}{\delta R(\mathbf{r})}
  +
  \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
  \frac{\delta n_2(\mathbf{r}')}{\delta R(\mathbf{r})}
  + \cdots
  \right) d\mathbf{r}'
\end{align}
There are \emph{many} such terms, and this could get expensive.

@@

Finally, we get an answer like
\begin{align}
  \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
  \int \mathbf{dr}' \left\{
  \left(
  \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
  + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
    n_3(\mathbf{r}'))^2}
  + \frac{1}{4\pi}\frac{
    \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
  }{
    (1-n_3(\mathbf{r}'))^3
  }
  \right) n(\mathbf{r}') \delta(|\mathbf{r}-\mathbf{r}'| - R)
  + \cdots
  \right\}
\end{align}
This comes out to quite a mess.  It wouldn't be infeasible to compute
this, but it wouldn't be efficient or pretty either.

@@

Unfortunately, knowing the average contact density on each hard sphere
doesn't really give us what we want for SAFT.  The trouble is that in
order to work out the fraction of association sites occupied, we need
to know the contact density \emph{at the point of contact}.
Fortunately, we don't need to know the contact density for a
particular sphere at each point of contact.  Just as $n_0(\mathbf{r})$
gives us the density of spheres that are contacting point $\mathbf{r}$
(see Figure~\ref{fig:n0}), we can use similar reasoning to work out the
contact density for these spheres.

To find the contact density at a given point of contact, we can't
simply change the radius for spheres touching that point, since that
would give us the pressure averaged over the entire surface of those
spheres.  Instead, we'll have to imagine extending a patch of those
spheres which is located at the point of interest, as illustrated in
Figure~\ref{fig:contact}.

@@

@@

We need to then find the change in Helmholtz free energy due to this
change at position $\mathbf{r}$:
\begin{align}
  \Delta A_{HS} &= \int \mathbf{dr}'
  \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}\Delta n_3(\mathbf{r}')
  +
  \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}\Delta n_2(\mathbf{r}')
  +
  \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} \cdot \Delta \mathbf{n}_{V2}(\mathbf{r}')
\end{align}

\end{widetext}

@@

>>>>

{
  \color{blue}We need to figure out a few bits of terminology.

  \begin{enumerate}
  \item What should we call the contact density between two hard
    spheres, to distinguish it from the contact density at any other
    hard surface.  One option would be to focus on the contact value
    of the correlation function, which would make it clear what we're
    talking about.
  \item How should we distinguish between the various definitions.  In
    particular, ``at the sphere'' vs. ``at contact''.  It'd be helpful
    to have a short way of describing these.
  \end{enumerate}
}

@@

\derivation{
  \begin{widetext}
}

@@

One reasonably simple question to ask is what the \emph{mean} contact
density of an inhomogeneous system is.  This comes down to simply
reproducing the derivation for a homogeneous system using the
FMT functional.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  \ncontact &= \frac{1}{N k_BT 4\pi (2R)^2} \frac{dA_{HS}}{dR}
\end{align}
While this should give the true mean contact density for any given
system, this is not particularly useful, as we have no way of knowing
the contact density on any given sphere.  However, the mean contact
density comes out as an integral, so we can plot the integrand, and
treat it as a somewhat ill-defined local contact density.

\derivation{

  To compute the mean contact density of a system, we need
  \begin{align}
    \frac{d A_{HS}}{d R} &=
    \frac{\partial A_{HS}}{\partial R} \\
    &+
    \int \left(
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
    \frac{d n_3(\mathbf{r}')}{d R}
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
    \frac{d n_2(\mathbf{r}')}{d R}
    + \cdots
    \right) d\mathbf{r}'
  \end{align}

@@

  \begin{align}
    \frac{d A_{HS}}{d R} &=
    \int \mathbf{dr}' \left(
    -\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3)
    + \cdots
    \right)
  \end{align}
}

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{4\pi (2R)^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  \ncontact(\xx) &= \frac{1}{k_BT 4\pi (2R)^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})}
\end{align}
After considerable math to do the above functional derivative, we find
a functional giving the contact density for spheres at any given
position.  This functional unavoidably involves more convolutions than
are present in the free energy itself, which makes it relatively
expensive.

\derivation{
  So to compute this we need
  \begin{align}
    \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
    \int \left(
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
    \frac{\delta n_3(\mathbf{r}')}{\delta R(\mathbf{r})}
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
    \frac{\delta n_2(\mathbf{r}')}{\delta R(\mathbf{r})}
    + \cdots
    \right) d\mathbf{r}'
  \end{align}
  There are \emph{many} such terms, and this could get expensive.

@@

  Finally, we get an answer like
  \begin{align}
    \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
    \int \mathbf{dr}' \left\{
    \left(
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{1}{4\pi}\frac{
      \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
    }{
      (1-n_3(\mathbf{r}'))^3
    }
    \right) n(\mathbf{r}') \delta(|\mathbf{r}-\mathbf{r}'| - R)
    + \cdots
    \right\}
  \end{align}
  This comes out to quite a mess.  It wouldn't be infeasible to compute
  this, but it wouldn't be efficient or pretty either.
}

@@

A final definition of the inhomogeneous contact density would be the
contact density for spheres \emph{touching at a given point}.  This
would be a symmetric measure, since both spheres involved are treated
in the same way.  In fact, the density of such spheres is just the
density $n_0(\xx)$ from FMT.  See Figure~\ref{fig:n0} for an
illustration $n_0$.  Since $n_0$ is the fundamental measure of sphere
surface contact, it would make sense to define a contact density that
would correspond.

To find the contact density at a given point of contact, we can't
simply change the radius for spheres touching that point, since that
would give us the pressure averaged over the entire surface of those
spheres.  Instead, we would have to imagine extending a patch of those
spheres which is located at the point of interest, as illustrated in
Figure~\ref{fig:contact}.  \textcolor{red}{I'm not sure how to go
  about doing this.  In the derivation section, I have some ideas
  jotted down, but they don't really work.}

Yu and Wu developed a functional that should correspond to the contact
density at a given contact
position~\cite{yu2002fmt-dft-inhomogeneous-associating}.  We will
compare both their functional as well as ours with simulation data.

@@

\derivation{

@@

  We need to then find the change in Helmholtz free energy due to this
  change at position $\mathbf{r}$:
  \begin{align}
    \Delta A_{HS} &= \int \mathbf{dr}'
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}\Delta n_3(\mathbf{r}')
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}\Delta n_2(\mathbf{r}')
    +
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} \cdot \Delta \mathbf{n}_{V2}(\mathbf{r}')
  \end{align}
}

\derivation{
  \end{widetext}
}

@@

\subsection{Hard spheres around a hard sphere (or rod)}

\textcolor{red}{
  It would also be interesting to study a system with opposite
  curvature.  Put a (larger) hard sphere in the hard-sphere fluid, and
  see what the contact density is around that hard sphere.  This
  touches on an important limiting case...
}
